Quotients of the magmatic operad:\\ lattice structures and convergent rewrite systems

### 2018-09-07 Introduction.tex
Passe introduction

This work is intended to study and collect the possible links between
the Tamari order and some quotients of the operad $\Mag$. In the long
run, we would like to study quotients $\Mag/_{\Congr}$ of $\Mag$ where
$\Congr$ is an operad congruence generated by equivalence classes of
trees of a fixed degree (that is, a fixed number of internal nodes). In
particular, we would like to know if the fact that $\Congr$ is generated
by equivalence classes of trees forming intervals of the Tamari order
implies algebraic properties for $\Mag/_{\Congr}$ (like the description
of orientations of its space of relations, nice bases, and Hilbert
series).
\medbreak

To explore this vast research area, we select to pursue in this paper
the following directions. First, we consider the very general set of the
quotients of $\Mag$ seen as an operad in the category of vector spaces.
We show that these operads form a lattice, wherein its partial order
relation is defined from the existence of operad morphisms
(Theorem~\ref{thm:lattice_structure_of_QMag}). We also provide a
Grassmann formula analog relating the Hilbert series of the operads of
the lattice together with their lower-bound and upper-bound
(Theorem~\ref{thm:Grassmann_formula_for_Hilbert_series_of_QMag}).
Besides, we focus on a special kind of quotients of $\Mag$, denoted by
$\CAs{\gamma}$, defined by equating the left and right comb binary trees
of a fixed degree $\gamma \geq 1$.
Observe that since $\CAs{2}=\As$, the operads $\CAs{\gamma}$ can be seen
as generalizations of $\As$. These operads are called comb
associative operads. For instance, $\CAs{3}$ is the operad describing
the category of the algebras equipped with a binary product $\Product$
subjected to the relation
\begin{equation}
    \left(\left(x_1 \Product x_2\right) \Product x_3\right) \Product x_4
    =
    x_1 \Product \left(x_2 \Product \left(x_3 \Product x_4\right)\right).
\end{equation}
We first provide general results about the
operads $\CAs{\gamma}$. In particular, we show that these operads form a
a lattice which embeds as a poset in the lattice of the quotients of
$\Mag$ aforementioned
(Theorems~\ref{thm:lattice_CAs} and~\ref{thm:inclusion_lattice_CAs}). We
focus in particular on the study of $\CAs{3}$. Observe that the
congruence of this operad is generated by an equivalence class of trees
which is not an interval for the Tamari order. As preliminary computer
experiments show, $\CAs{3}$ has oscillating first dimensions
(see~\eqref{equ:dimensions_CAs_3}), what is rather unusual among all
known operads. We provide a convergent orientation of the space of
relations of $\CAs{3}$ (Theorem~\ref{thm:convergent_rewrite_rule_CAs_3}),
a description of a basis of the operad in terms of normal forms, and
prove that its Hilbert series is rational. For all these, we use rewrite
systems on trees~\cite{BN98} and the Buchberger algorithm for
operads~\cite{DK10}. We expose some experimental results obtained with
the help of the computer for some operads $\CAs{\gamma}$ with
$\gamma \geq 4$. Finally, we continue the
investigation of the quotients of $\Mag$ by regarding the quotients of
$\Mag$ obtained by equating two trees of degree $3$. This leads to ten
quotient operads of $\Mag$. We provide for some of these combinatorial
realizations, mostly in terms of integer compositions.
\medbreak

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Acknoledgements}
The authors wish to thank Maxime Lucas for helpfull discussions and
Vladimir Dotsenko for his marks of interest and his bibliographic
suggestions.
\medbreak

>>>>

This work is intended to study and collect the possible links between
the Tamari order and some quotients of the operad $\Mag$. In the long
run, the goal is to study quotients $\Mag/_{\Congr}$ of $\Mag$ where
$\Congr$ is an operad congruence generated by equivalence classes of
trees of a fixed degree (that is, a fixed number of internal nodes). In
particular, we would like to know if the fact that $\Congr$ is generated
by equivalence classes of trees forming intervals of the Tamari order
implies algebraic properties for $\Mag/_{\Congr}$ (like the description
of orientations of its space of relations, of nice bases, and of Hilbert
series).
\medbreak

To explore this vast research area, we select to pursue in this paper
the following directions. First, we consider the very general set of the
quotients of $\Mag$ seen as an operad in the category of vector spaces.
We show that this set of operads forms a lattice, wherein its partial
order relation is defined from the existence of operad morphisms
(Theorem~\ref{thm:lattice_structure_of_QMag}). We also provide a
Grassmann formula analog relating the Hilbert series of the operads of
the lattice together with their lower-bound and upper-bound
(Theorem~\ref{thm:Grassmann_formula_for_Hilbert_series_of_QMag}).
Besides, we focus on a special kind of quotients of $\Mag$, denoted by
$\CAs{\gamma}$, defined by equating the left and right comb binary trees
of a fixed degree $\gamma \geq 1$. Observe that since $\CAs{2}=\As$, the
operads $\CAs{\gamma}$ can be seen as generalizations of $\As$. These
operads are called comb associative operads. For instance, $\CAs{3}$ is
the operad describing the category of the algebras equipped with a
binary product $\Product$ subjected to the relation
\begin{equation}
    \left(\left(x_1 \Product x_2\right) \Product x_3\right) \Product x_4
    =
    x_1 \Product \left(x_2 \Product \left(x_3 \Product x_4\right)\right).
\end{equation}
We first provide general results about the operads $\CAs{\gamma}$. In
particular, we show that the set of these operads forms a lattice which
embeds as a poset in the lattice of the quotients of $\Mag$
aforementioned (Theorems~\ref{thm:lattice_CAs}
and~\ref{thm:inclusion_lattice_CAs}). We focus in particular on the
study of $\CAs{3}$. Observe that the congruence defining this operad is
generated by an equivalence class of trees which is not an interval for
the Tamari order. As preliminary computer experiments show, $\CAs{3}$
has oscillating first dimensions (see~\eqref{equ:dimensions_CAs_3}),
what is rather unusual among all known operads. We provide a convergent
orientation of the space of relations of $\CAs{3}$
(Theorem~\ref{thm:convergent_rewrite_rule_CAs_3}), a description of a
basis of the operad in terms of normal forms, and prove that its Hilbert
series is rational. For all these, we use rewrite systems on
trees~\cite{BN98} and the Buchberger algorithm for operads~\cite{DK10}.
We expose some experimental results obtained with the help of the
computer for some operads $\CAs{\gamma}$ with $\gamma \geq 4$. Finally,
we continue the investigation of the quotients of $\Mag$ by regarding
the quotients of $\Mag$ obtained by equating two trees of degree $3$.
This leads to ten quotient operads of $\Mag$. We provide for some of
these combinatorial realizations, mostly in terms of integer
compositions.
\medbreak

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Acknoledgements}
The authors wish to thank Maxime Lucas for helpful discussions and
Vladimir Dotsenko for his marks of interest and his bibliographic
suggestions.
\medbreak
